---
title: Inverse planning — desire inference, alternatives shown (data analysis)
author: Alicia Chen
date: 2026-01-15
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("food_inv_desire_intimacy_alt")
```

## Results

This experiment presents participants with scenarios where they observe an action taken by one person toward another. Participants rate their belief about the probability that the actors have high motivation (want the food) before (prior) and after (posterior) observing the action. The experiment manipulates:
- **Action**: Which action was taken (action 0-3, varying in saliva-transfer risk)
- **Intimacy**: The relationship closeness between the actors (0, 50, 75, or 100)

## Load human data

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "food_inv_desire_intimacy_alt", "main_trials_long.csv")) |>
  mutate(
    action = str_replace(action_condition, "action_", ""),
    intimacy = factor(intimacy),
    p_high_reward = response / 100,  # Convert to probability scale
    stage = factor(stage, levels = c("prior", "posterior")),
    model = "Humans"
  ) |>
  select(-action_condition, -response)

# Calculate belief updates
df <- calculate_belief_update(df, "p_high_reward")
```

## Human data: Prior vs posterior ratings

### Distribution of ratings

```{r}
#| fig-width: 12
#| fig-asp: 0.45

ggplot(df, aes(x = p_high_reward, fill = action)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(y = "Density", x = "P(high motivation)", fill = "Action") +
  theme(legend.position = "right") +
  facet_grid(stage ~ intimacy, labeller = labeller(intimacy = label_both, stage = label_value))
```

### Summary statistics

```{r}
df_summary <- df |>
  group_by(intimacy, action, stage, model) |>
  tidyboot_mean(p_high_reward)
```

### Prior vs posterior by action and intimacy

```{r}
#| fig-width: 12
#| fig-asp: 0.7

ggplot() +
  geom_violin(
    data = df,
    aes(x = stage, y = p_high_reward, fill = action),
    position = position_dodge(width = 0.8),
    alpha = 0.5,
    width = 0.7
  ) +
  geom_point(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 3,
    alpha = PLOT_ALPHA
  ) +
  geom_errorbar(
    data = df_summary,
    aes(x = stage, ymin = ci_lower, ymax = ci_upper, color = action, group = action),
    position = position_dodge(width = 0.8),
    width = 0.1,
    size = 1
  ) +
  geom_line(
    data = df_summary,
    aes(x = stage, y = empirical_stat, color = action, group = action),
    position = position_dodge(width = 0.8),
    size = 1
  ) +
  facet_grid(intimacy ~ action, labeller = label_both) +
  scale_fill_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  scale_color_viridis_d(option = "plasma", begin = 0, end = 0.85) +
  labs(
    title = "P(high motivation) before and after observing action",
    y = "P(high motivation)",
    x = "Stage",
    fill = "Observed action",
    color = "Observed action"
  )
```

### Belief updates

```{r}
df_belief_update_summary <- df |>
  filter(stage == "posterior") |>
  group_by(intimacy, action, model) |>
  tidyboot_mean(belief_update)
```

## Model comparison


### Fitted parameters

```{r}
df_fit_results <- read_csv(here("model", "outputs", "food_inv_desire_intimacy_alt", "fit_results.csv")) |>
  filter(experiment == "food_inv_desire_intimacy_alt") |>
  mutate(
    model_label = relabel_model(model)
  ) |>
  filter(!is.na(model_label)) |>
  select(model_label, alpha_observer, sse, mse, n_params)

df_fit_results |>
  knitr::kable(
    digits = 3,
    caption = "Fitted observer parameters for desire inference",
    col.names = c("Model", "α_observer", "SSE", "MSE", "n_params")
  )
```

### AIC comparison

Following Burnham and Anderson (2002), ΔAIC values of 0–2 indicate essentially equivalent models, 4–7 indicate considerably less support for the higher-AIC model, and values greater than 10 indicate essentially no support for the higher-AIC model.

```{r}
kable_aic_table(df_fit_results, loss_col = "sse", loss_col_label = "SSE")
```

### Load model predictions

```{r}
df_model <- read_csv(here("model", "outputs", "food_inv_desire_intimacy_alt", "cv_preds_summary.csv")) |>
  rename(intimacy = intimacy_condition) |>
  mutate(
    action = as.character(action),
    intimacy = factor(intimacy),
    # Model predictions are P(high desire) in percentage
    p_high_reward = p_high_reward / 100,
    # Belief update from uniform prior (50%)
    belief_update = p_high_reward - 0.5,
    model = relabel_model(model)
  ) |>
  filter(!is.na(model))

MODEL_LEVELS <- c("Base model", "Discomfort-only", "Full model", "Humans")
```

### Combine human and model data

```{r}
df_combined <- bind_rows(
  df_belief_update_summary |>
    mutate(belief_update = empirical_stat) |>
    select(intimacy, action, model, belief_update, ci_lower, ci_upper),
  df_model |>
    group_by(intimacy, action, model) |>
    summarize(belief_update = mean(belief_update), .groups = "drop") |>
    mutate(ci_lower = NA_real_, ci_upper = NA_real_)
) |>
  mutate(
    model = factor(model, levels = MODEL_LEVELS),
    action = factor(action)
  )
```

### Belief updates: Model vs human comparison

```{r}
#| fig-width: 12
#| fig-asp: 0.3

df_combined |>
  ggplot(aes(x = action, y = belief_update, group = interaction(model, intimacy), fill = intimacy)) +
  geom_crossbar(
    aes(y = 0, ymin = -0.015, ymax = 0.015),
    position = position_dodge(width = 0.9),
    width = 0.9,
    color = NA
  ) +
  geom_col(position = position_dodge(width = 0.9), alpha = PLOT_ALPHA) +
  geom_errorbar(
    aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.9), size = 1, width = 0, color = "black", alpha = 0.8
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_fill_intimacy() +
  labs(y = "Belief update", x = "Observed action", fill = "Intimacy") +
  theme(legend.position = "right") +
  facet_wrap(~model, nrow = 1)

ggsave(here("figures", "food_inv_desire_intimacy_alt_belief_main.pdf"), device = cairo_pdf)
```

