---
title: Forward planning (data analysis)
author: Alicia Chen
date: 2026-01-03
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    page-layout: full
    search: false
knitr:
  opts_chunk:
    dev: "ragg_png"
execute:
  warning: false 
---

```{r setup}
source(here::here("analysis", "utils.R"))
setup_analysis()
```

## Demographics

```{r}
report_demographics("food_forw_intimacy_desire")
```

## Results

This experiment manipulates how much two people want the food (`low` vs. `high` desire), and the relationship between the two people (0, 50, 75, or 100 on a 100 point scale between 'maximally intimate' and 'maximally formal'). 

Load csv, calculate means

```{r}
# main_trials_long.csv is generated by analysis/json_to_csv.py
# It excludes participants who failed attention check or got 0 on memory check
df <- read_csv(here("data", "food_forw_intimacy_desire", "main_trials_long.csv")) |>
  mutate(
    intimacy = factor(intimacy),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS),
    model = "humans"
  )

df_summary <- df |>
  group_by(intimacy, motivation, action) |>
  tidyboot_mean(p_action)
```

## Model comparison


### Load model predictions

Model predictions are the **out-of-sample predictions** from `model/outputs/food_forw_intimacy_desire/cv_preds.csv`: for each trial, the prediction was made by a model fit on the other 15 scenarios, never seeing that trial's scenario during training. This is what we report in the paper.

```{r}
df_cv <- read_csv(here("model", "outputs", "food_forw_intimacy_desire", "cv_preds.csv"), show_col_types = FALSE)

# Humans (take from one variant — all three variants share the same p_action column)
df_humans_cv <- df_cv |>
  filter(variant == "full") |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, motivation, action,
    p_action, model = "Humans"
  )

# Out-of-sample model predictions, one row per trial per variant
df_model_cv <- df_cv |>
  mutate(
    model = relabel_model(variant)
  ) |>
  transmute(
    subject_id, scenario_label = held_out_scenario,
    intimacy, motivation, action,
    p_action = p_action_pred, model
  )

df_preds <- bind_rows(df_humans_cv, df_model_cv) |>
  mutate(
    intimacy = factor(intimacy),
    model = factor(model, levels = c(
      "Base model", "Discomfort-only", "Full model", "Humans"
    )),
    motivation = factor(motivation, levels = c("low", "high"), labels = MOTIVATION_LEVELS)
  )

df_preds_summary <- df_preds |>
  group_by(intimacy, motivation, action, model) |>
  tidyboot_mean(p_action) |>
  ungroup()
```

### Model predictions vs human data

```{r}
#| fig-width: 12
#| fig-asp: 0.43

df_preds_summary |>
  ggplot(aes(x = action, y = empirical_stat, color = intimacy, shape = factor(action), group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 4, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), size = 1.2, width = 0.5
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1.2) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(y = "P(Action)", x = "Action", color = "Intimacy", shape = "Action") +
  facet_grid(motivation ~ model)

ggsave(here("figures", "food_forw_intimacy_desire_comparison.pdf"),
         device = cairo_pdf)
```

### Model-human correlation

```{r}
# Bootstrap function for correlation CI (handles NAs)
boot_cor <- function(x, y, n_boot = 1000) {
  # Remove NAs
  complete <- complete.cases(x, y)
  x <- x[complete]
  y <- y[complete]
  n <- length(x)
  if (n < 3) return(list(r = NA_real_, ci_lower = NA_real_, ci_upper = NA_real_))
  boot_rs <- replicate(n_boot, {
    idx <- sample(n, replace = TRUE)
    cor(x[idx], y[idx])
  })
  list(
    r = cor(x, y),
    ci_lower = as.numeric(quantile(boot_rs, 0.025, na.rm = TRUE)),
    ci_upper = as.numeric(quantile(boot_rs, 0.975, na.rm = TRUE))
  )
}

# Get human data for joining
df_humans <- df_preds_summary |>
  filter(model == "Humans") |>
  select(intimacy, motivation, action,
         Humans = empirical_stat,
         Humans_ci_lower = ci_lower,
         Humans_ci_upper = ci_upper)

# Create correlation data by joining model predictions with human data
df_corr_all <- df_preds_summary |>
  filter(model != "Humans") |>
  select(intimacy, motivation, action, model, model_pred = empirical_stat) |>
  left_join(df_humans, by = c("intimacy", "motivation", "action")) |>
  filter(!is.na(model_pred), !is.na(Humans)) |>
  mutate(action = factor(action))

# Debug: check what models exist
cat("Models in df_corr_all:", unique(as.character(df_corr_all$model)), "\n")

# Use all models for correlation
df_corr <- df_corr_all |>
  mutate(model = droplevels(model))

cat("Rows in df_corr:", nrow(df_corr), "\n")
```

```{r}
#| label: fig-food-forw-intimacy-desire-correlation
#| fig-width: 10
#| fig-asp: 0.4


# Read the bootstrap r/CI computed by scripts/generate_values_tex.py so the
# plot labels exactly match the values.tex table (avoids cross-language
# bootstrap RNG drift).
correlations <- read_csv(
  here("model", "outputs", "food_forw_intimacy_desire", "cv_correlation.csv"),
  show_col_types = FALSE
) |>
  mutate(
    model = factor(relabel_model(variant), levels = MODEL_LEVELS),
    label = paste0(
      "r = ", sprintf("%.2f", r),
      " (", sprintf("%.2f", ci_lower), ", ", sprintf("%.2f", ci_upper), ")"
    )
  ) |>
  select(model, r, ci_lower, ci_upper, label)

ggplot(df_corr,
                        aes(
                          x = model_pred,
                          y = Humans,
                          color = intimacy,
                          shape = action
                        )) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "lightgray"
  ) +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08,
    vjust = 1.5,
    size = 4.5,
    fill = "white",
    alpha = 0.5,
    label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(
    x = "Model",
    y = "Humans",
    color = "Intimacy",
    shape = "Action"
  ) +
  facet_wrap( ~ model, nrow = 1) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)

ggsave(here("figures", "food_forw_intimacy_desire_correlation.pdf"),
         device = cairo_pdf)

# p_correlation
#
# save_figure(p_correlation, "food_forw_intimacy_desire_correlation.pdf")
```

Plot motivation on separate panels:

```{r fig.width=12, fig.height=6}
correlations_motiv <- format_correlation_labels(df_corr, model_pred, Humans,
                                                 group_vars = c("model", "motivation"))

ggplot(df_corr, aes(x = model_pred, y = Humans, color = intimacy, shape = action)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "lightgray") +
  geom_errorbar(aes(ymin = Humans_ci_lower, ymax = Humans_ci_upper), width = 0.02) +
  geom_point(size = 3, alpha = PLOT_ALPHA) +
  geom_label(
    data = correlations,
    aes(x = -Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = -0.08,
    vjust = 1.5,
    size = 4.5,
    fill = "white",
    alpha = 0.5,
    label.size = 0
  ) +
  scale_color_intimacy() +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Model", y = "Humans", color = "Intimacy", shape = "Action") +
  facet_grid(motivation~model) +
  coord_fixed_symmetric(df_corr$model_pred, df_corr$Humans)
```


Plot each scenario separately


```{r}
#| label: fig-food-forw-intimacy-desire-scenario-summary
#| fig-width: 23 
#| fig-asp: 0.18


df_summary_scenario <- df |>
  group_by(scenario_label, intimacy, motivation, action) |>
  tidyboot_mean(p_action)

ggplot(df_summary_scenario, aes(x = action, y = empirical_stat, color = intimacy, group = intimacy)) +
  geom_point(position = position_dodge(width = 0.2), size = 3, alpha = PLOT_ALPHA) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
    position = position_dodge(width = 0.2), width = 0.4
  ) +
  geom_line(position = position_dodge(width = 0.2), linewidth = 1) +
  scale_color_intimacy() +
  labs(y = "P(action)", x = "Action", color = "Intimacy") +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(motivation~scenario_label, labeller = labeller(scenario_label = label_wrap_gen(15)))
```


### Model fit results

```{r}
fit_results <- read_csv(here("model", "outputs", "food_forw_intimacy_desire", "fit_results.csv"))
fit_results |>
  select(model, nll, n_params, starts_with("param_")) |>
  knitr::kable(digits = 3, caption = "Model fitting results")
```

The full risk model provides the best fit (lowest NLL), followed by the discomfort-only ablation and the base model.

### AIC/BIC comparison

```{r}
fit_results <- add_aic_bic(fit_results, n_obs = nrow(df))

fit_results |>
  select(model, nll, n_params, AIC, BIC) |>
  arrange(AIC) |>
  knitr::kable(digits = 2, caption = "Model comparison (lower AIC/BIC is better)")
```

```{r}
# Compute differences from best model (full)
best_aic <- fit_results$AIC[fit_results$model == "full"]
best_bic <- fit_results$BIC[fit_results$model == "full"]

fit_results |>
  mutate(
    delta_AIC = AIC - best_aic,
    delta_BIC = BIC - best_bic
  ) |>
  select(model, delta_AIC, delta_BIC) |>
  knitr::kable(digits = 2, caption = "Difference from best model (Full)")
```

## Residual diagnostics

We claimed humans vary more across scenarios than the Full model does. These diagnostics quantify that claim and surface which scenarios / actions drive the unexplained variance.

### Setup

```{r}
# Per-cell means for Humans and Full model
df_cell <- df_preds |>
  filter(model %in% c("Humans", "Full model")) |>
  group_by(scenario_label, intimacy, motivation, action, model) |>
  summarize(p_action = mean(p_action), .groups = "drop") |>
  pivot_wider(names_from = model, values_from = p_action) |>
  rename(full_model = `Full model`, humans = Humans) |>
  mutate(residual = humans - full_model)
```

### Cross-scenario dispersion: humans vs Full model

For each (intimacy, motivation, action) condition, compute the SD across scenarios of P(action). If the model captures scenario-level variance well, the blue and red bars should match. Gaps = variance the model misses.

```{r}
#| fig-width: 12
#| fig-asp: 0.45

df_dispersion <- df_cell |>
  group_by(intimacy, motivation, action) |>
  summarize(
    Humans = sd(humans),
    `Full model` = sd(full_model),
    .groups = "drop"
  ) |>
  pivot_longer(c(Humans, `Full model`), names_to = "source", values_to = "sd") |>
  mutate(source = factor(source, levels = c("Full model", "Humans")))

ggplot(df_dispersion, aes(x = factor(action), y = sd, fill = source)) +
  geom_col(position = position_dodge(width = 0.8), alpha = PLOT_ALPHA, width = 0.75) +
  scale_fill_manual(values = c("Full model" = "#6b8cae", "Humans" = "#b4584d")) +
  labs(x = "Action", y = "SD across scenarios of P(action)", fill = NULL) +
  facet_grid(motivation ~ intimacy, labeller = labeller(intimacy = label_both)) +
  theme(legend.position = "top")
```

### Per-scenario fit quality

Each scenario has 32 cells (4 intimacy × 2 motivation × 4 action). Compute Pearson r between Humans and Full model over those cells. Low r = scenario the model systematically misreads.

```{r}
#| fig-width: 8
#| fig-asp: 0.7

df_scenario_fit <- df_cell |>
  group_by(scenario_label) |>
  summarize(
    r = cor(humans, full_model),
    rmse = sqrt(mean(residual^2)),
    .groups = "drop"
  ) |>
  arrange(r)

ggplot(df_scenario_fit, aes(x = reorder(scenario_label, r), y = r)) +
  geom_col(fill = "#7A4A5A", alpha = PLOT_ALPHA) +
  geom_text(aes(label = sprintf("%.2f", r)), hjust = -0.15, size = 3.5) +
  coord_flip(ylim = c(0, 1.05)) +
  labs(x = "Scenario", y = "Pearson r (Humans vs Full model)")
```

### Mean residual per (scenario, action)

Heatmap of mean signed residual (humans − full model), averaged over intimacy × motivation. Red = model under-predicts P(action) there; blue = over-predicts. Scenarios are ordered by mean |residual|.

```{r}
#| fig-width: 7
#| fig-asp: 1.1

df_cell_action <- df_cell |>
  group_by(scenario_label, action) |>
  summarize(mean_residual = mean(residual), .groups = "drop")

scenario_order <- df_cell_action |>
  group_by(scenario_label) |>
  summarize(mabs = mean(abs(mean_residual))) |>
  arrange(mabs) |>
  pull(scenario_label)

df_cell_action |>
  mutate(
    scenario_label = factor(scenario_label, levels = scenario_order),
    action = factor(action)
  ) |>
  ggplot(aes(x = action, y = scenario_label, fill = mean_residual)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%+.2f", mean_residual)), size = 3) +
  scale_fill_gradient2(low = "#3b4cc0", mid = "white", high = "#b40426",
                       midpoint = 0, limits = c(-1, 1) * max(abs(df_cell_action$mean_residual))) +
  labs(x = "Action", y = "Scenario", fill = "Residual\n(H − M)") +
  theme(panel.grid = element_blank())
```

